---
title: "Homework 1"
subtitle: "PSTAT 131/231, Spring 2019"
author: "Eduardo Escoto (Perm: 7611817) / JunRong Ma (Perm: 1234567) / (PSTAT 131)"
date: "__Due on April 20, 2019 at 11:59 pm__"
graphics: yes
output: 
  pdf_document
knit: (function(output_file, encoding){rmarkdown::render(output_file, output_format = "pdf_document", output_dir = "../output")})
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(reshape2)
indent1 = '    '
indent2 = paste(rep(indent1, 2), collapse = '')
indent3 = paste(rep(indent1, 3), collapse = '')
r = function(x, digits = 2) {
  round(x, digits = digits)
}
```

```{r setting_chunk_options, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = FALSE,
  fig.width = 5,
  fig.height = 5,
  fig.align = 'center'
)
```

__Note:__ If you are working with a partner, please submit only one homework per group with both names and whether you are taking the course for graduate credit or not.  Submit your Rmarkdown (.Rmd) and the compiled pdf on Gauchospace.
 
---------------------------------

**Predicting Algae Blooms**

__*Background*__ High concentrations of certain harmful algae in rivers
constitute a serious ecological problem with a strong impact not only on river
lifeforms, but also on water quality. Being able to monitor and perform an
early forecast of algae blooms is essential to improving the quality of rivers.

With the goal of addressing this prediction problem, several water samples were
collected in different European rivers at different times during a period of
approximately 1 year. For each water sample, different chemical properties were
measured as well as the frequency of occurrence of seven harmful algae. Some
other characteristics of the water collection process were also stored, such as
the season of the year, the river size, and the river speed.

__*Goal*__ We want to understand how these frequencies are related to certain
chemical attributes of water samples as well as other characteristics of the
samples (like season of the year, type of river, etc.)
    
__*Data Description*__ The data set consists of data for 200 water samples and
each observation in the available datasets is in effect an aggregation of
several water samples collected from the same river over a period of 3 months,
during the same season of the year. Each observation contains information on 11
variables. Three of these variables are nominal and describe the season of the
year when the water samples to be aggregated were collected, as well as the
size and speed of the river in question. The eight remaining variables are
values of different chemical parameters measured in the water samples forming
the aggregation, namely: Maximum pH value, Minimum value of $O_2$ (oxygen),
Mean value of Cl (chloride), Mean value of $NO_3^-$ (nitrates), Mean value of
$NH_4^+$ (ammonium), Mean of $PO^{3}_4$ (orthophosphate), Mean of total $PO_4$
(phosphate) and Mean of chlorophyll.

Associated with each of these parameters are seven frequency numbers of
different harmful algae found in the respective water samples. No information
is given regarding the names of the algae that were identified.
    
We can start the analysis by loading into R the data from the "algaeBloom.txt"
file (the training data, i.e. the data that will be used to obtain the
predictive models). To read the data from the file it is sufficient to issue
the following command:

```{r load, message=F, warning=F, results="hide"}
algae <- read_table2("../algae_data/algaeBloom.txt", col_names=
                      c('season','size','speed','mxPH','mnO2','Cl','NO3','NH4',
                        'oPO4','PO4','Chla','a1','a2','a3','a4','a5','a6','a7'), 
                      na="XXXXXXX")

glimpse(algae)
```

1. __*Descriptive summary statistics*__ Given the lack of further information
   on the problem domain, it is wise to investigate some of the statistical
   properties of the data, so as to get a better grasp of the problem. It is
   always a good idea to start our analysis with some kind of exploratory data
   analysis. A first idea of the statistical properties of the data can be
   obtained through a summary of its descriptive statistics. 
    
    #. Count the number of observations in each season using `summarize()` in
    `dplyr`. 
    
        #. The numbers of observations for each season are calculated below and displayed in the `n_obs` column in the tibble below.
    
```{r 1.a) Number of observations, indent = indent3}
number_of_obs_by_season <- algae %>% 
  # Group observations by season
  group_by(season) %>%
  # Counts the number of observations in the groups
  summarize(n_obs = n()) 

number_of_obs_by_season
```

    #. Are there missing values? Calculate the mean and variance of each
    chemical (Ignore $a_1$ through $a_7$). What do you notice about the
    magnitude of the two quantities for different chemicals? 
    
        #. There are indeed missing values! In the tibble below there is a preview of missing values.
      
```{r 1.b) Missing Values, indent = indent3}
algae_missing_values <- algae %>%
  # Filters on all columns on if any column contains an NA
  filter_all(any_vars(is.na(.)))

algae_missing_values
```
      
        #. The Means and Variances for each chemical are calculated below and displayed in the tibble.
      
```{r 1.b) Mean and Variance of each chemical, indent = indent3}
m_v_by_chemical <- algae %>%
  # Selects All chemicals
  select(mnO2:Chla) %>%
  # Groups by key THEN removes NA's
  gather(chemical,factor_key = TRUE) %>% group_by(chemical) %>% na.omit() %>%
  # Calculates mean and variance
  summarize_all(list(mean = ~ mean(.),
                     variance = ~ var(.)))
m_v_by_chemical
```

        #. As you can see in the tibble above, the magnitudes of thes Means for each chemical seem to vary highly and they all have large Variances with high magnitudes. That means that there is alot of spread between all of the data points for each chemical, except for $\text{mnO}2$ which has a variance of $`r m_v_by_chemical[1,3]`$ which is relatively small compared to the variance of the other chemicals.

    #. Mean and Variance is one measure of central tendency and spread of data.
    Median and Median Absolute Deviation are alternative measures of central
    tendency and spread. 

        For a univariate data set $X_1, X_2, ..., X_n$, the Median Absolute Deviation (MAD) is defined as the median of the absolute deviations from the data's median: $$\text{MAD}=\text{median} (|X_i-\text{median}(X)|)$$
        Compute median and MAD of each chemical and compare the two sets of quantities (i.e., mean & variance vs. median & MAD). What do you notice? 
        
```{r 1.c) MAD of each chemical, indent = indent2}
med_mad_by_chemical <- algae %>%
  # Selects only the chemicals
  select(mnO2:Chla) %>%
  # Removes the NA's AFTER grouping
  gather(chemical,factor_key = TRUE) %>% group_by(chemical) %>% na.omit() %>%
  # Calculates the MAD and Medians
  summarize_all(list(median = ~ median(.),
                     mad = ~ mad(., constant = 1, low = F, high = F)))

med_mad_by_chemical
```

        #. Between *Mean* and *Median* the values tend to be pretty close to eachother, except in the case of $\text{NH}4$ where the mean is about 400 more than the median. 
        
        #. The more noticable difference is the difference between *MAD* and *Variance.* Where the MAD is significantly lower compared to the Variance for the same chemical. A prime example of this is again, $\text{NH}4$ where the MAD is $`r med_mad_by_chemical[4,3]`$ which is 4 powers of 10 lower than the Variance of $`r m_v_by_chemical[4,3]`$. 
        
        #. These differences between Mean and Medain and MAD and Variacne might tell us that most values tend to be close together but there are some values for each chemical that are very large that are skewing the Means and Variances.


2. __*Data visualization*__ Most of the time, the information in the data set is also well captured graphically. Histogram, scatter plot, boxplot, Q-Q plot are frequently used tools for data visualization. Use ggplot for all of these visualizations. 
    
    #. Produce a histogram of $mxPH$ with the title 'Histogram of mxPH' based on algae data set. Use an appropriate argument to show the probability instead of the frequency as the vertical axis. (Hint: look at the examples in the help file for function `geom_histogram()`). Is the distribution skewed? 
        
    #. Add a density curve using `geom_density()` and rug plots using `geom_rug()` to above histogram. 
  
    #. Create a boxplot with the title 'A conditioned Boxplot of Algal $a_1$' for $a_1$ grouped by $size$. (Refer to help page for `geom_boxplot()`). 
  
    #. Are there any outliers for $NO3$ and $NH4$? How many observations would you consider as outliers? How did you arrive at this conclusion? 

    #. Compare mean & variance vs. median & MAD for $NO3$ and $NH4$. What do you notice? Can you conclude which set of measures is more robust when outliers are present? 
    
        
----------------------------------------------
    
**Predicting Algae Blooms**

Some water samples contained unknown values in
several chemicals. Missing data are very common in real-world problems, and
may prevent the use of certain data mining techniques that are not able to
handle missing values. 

In this homework, we are going to introduce various ways to deal with missing
values. After all the missing values have been taken care of, we will build a
model to investigate the relationship between the variable `a1` and other 11
predictors (`season`, `size`, `speed`, `mxPH`, `mnO2`, `Cl`, `NO3`, `NH4`,
`oPO4`, `PO4`, `Chla`) utilizing cross-validation in the next
problem.

**_Dealing with missing values_**

3.     
    #. How many observations contain missing values? How many missing values
    are there in each variable? 

```{r 3a) Observations With Missing Values, indent = indent2}
n_missing_obs <- algae %>% 
  filter_all(any_vars(is.na(.))) %>% nrow()

n_missing_obs
```

        #. There are $`r n_missing_obs`$ observations with missing values.

```{r 3a) Missing Values in each variable, indent = indent2}
n_missing_obs_by_variable <- algae %>% 
  gather(variable,factor_key = TRUE) %>% group_by(variable) %>% 
  summarize_all(list(n_missing_vals = ~ sum(is.na(.))))

n_missing_obs_by_variable
```

        #. This tibble displays the number of missing values per variable.

    #. **Removing observations with missing values**: use `filter()` function
    in `dplyr` package to observations with any missing value, and save the
    resulting dataset (without missing values) as `algae.del`. Report how many
    observations are in `algae.del`.
    
        Hint: `complete.cases()` may be useful.

```{r 3b) Missing Values Removed, indent = indent2}
algae.del <- algae %>%
  filter(complete.cases(.))

nrow(algae.del)
```

        #. There are $`r nrow(algae.del)`$ observations in `algae.del`

    #. \label{imputation} **Imputing unknowns with measures of central
    tendency**: the simplest and fastest way of filling in (imputing) missing
    values is to use some measures of central tendency such as mean, median and
    mode.
        
        Use `mutate_at()` and `ifelse()` in `dplyr` to fill in missing values
        for each chemical with its median, and save the imputed dataset as
        `algae.med`. Report the number of observations in `algae.med`.  Display
        the values of each chemical for the $48^{th}$, $62^{th}$ and $199^{th}$
        obsevation in `algae.med`. 

```{r 3c) Imputing values with medians, indent = indent2}
# Gets Columns with missing data into vector
missing_keys <- algae %>% 
  gather(factor_key = TRUE) %>% group_by(key) %>%
  filter_all(any_vars(is.na(.))) %>% tally() %>%
  pull(key)

# Imputes at all values in columns that have missing data
algae.med <- algae %>%
   mutate_at(.vars = vars(missing_keys),
             .funs = list(~ ifelse(is.na(.), median(., na.rm=TRUE), .)))
```

```{r 3c) print at, indent = indent2}
chemicals_algae.med = algae.med %>% select(mnO2:Chla)

displ_chem_imp <- rbind(chemicals_algae.med[48, ],
                        chemicals_algae.med[62, ],
                        chemicals_algae.med[199, ])
displ_chem_imp <- tribble(~ index, 48, 62, 199) %>% cbind(displ_chem_imp)

displ_chem_imp 
```

        #. The Tibble above shows the Imputed Values at the index requested.

        This simple strategy, although extremely fast and thus appealing for
        large datasets, imputed values may have large bias that can influence
        our model fitting. An alternative for decreasing bias of imputed values
        is to use relationships between variables.
        
    #. **Imputing unknowns using correlations**: another way to impute missing
    values is to use correlation with another variable. For a highly
    correlated pair of variables, we can fill in the unknown values by
    predicting one based on the other with a simple linear regression model,
    provided the two variables are not both unknown. 
    
        Compute pairwise correlation between the continuous (chemical) variables. 
        
```{r 3.d) pairwise corr, indent = indent2}
algae.pairwise_corr <- algae %>% select(mnO2:Chla) %>% 
  cor(use = 'complete.obs') %>% as_tibble(rownames = "id")

algae.pairwise_corr
```

        #. The Pairwise Correlations are computed and displayed above.

        Then, fill in the missing value for `PO4` based on `oPO4` in the
        $28^{th}$ observation. What is the value you obtain? 
        
        Hint: use `lm()` and `predict()` function.

```{r 3e) Imputing w corr, indent = indent2}
PO4.lm = lm(PO4 ~ oPO4, algae)

algae.PO4.pred <- algae %>%
  mutate(PO4 = ifelse(is.na(PO4), 
                      predict(PO4.lm, newdata = data.frame(oPO4)), PO4))

algae.PO4.pred[28,"PO4"]
```

        #. Above, the PO4 for the 28th observation is calculated using a linear model based on oPO4 due to the high pairwise correlation between them. In the 28th observation we have the value of $\text{oPO4} = `r algae.PO4.pred[28,"oPO4"]`$ which gives us a predicted value of $\text{PO4} = `r algae.PO4.pred[28,"PO4"]`$.

    #. **Questioning missing data assumptions**:  When might imputation using only the observed data lead you to incorrect conclusions?  In a couple of sentences, describe a scenario in which the imputed values of the chemical abundances in the algae data  (imputed using either the median or correlation method) might be a poor substitute for the true missing values.  Hint: look at the example from lecture 2.  
        
        #. Imputation might cause us to have incorrect conclusions becuase of relying too heavily on the observed data only. Lets say that in a scenario of the algae data we have new data that is far from the prediction based on oPO4 or far from the medians of each chemical, then we will have very high test error, and a model that is too overfitted on the training data.

        #. *In the context of Correlation Method:* In Lecture 2 we learned about *Survivorship Bias*. Many datesets have Survivorship Bias where the data that we have is insufficient in telling us about a certain variable. In the context of the algae data, using oPO4 to impute values of PO4 might be inducing Survivorship bias because oPO4 might actually not be sufficient in predicting PO4. Leading us to have high test error again
        
        #. *In the context of the Median Method:* this might lead us to the wrong conclusions by introducing alot of Bias because we are using only the observed data to impute. New data might be very far from the Median causing us to have values that have high test error.
        
        
**_Estimating the Test Error with Cross Validation (CV)_**
    
Using `algae.med` dataset obtained in \eqref{imputation}, we will build a linear regression model
to predict the levels of algae type `a1` based on 11 variables (`season`, `size`, `speed`, `mxPH`,
`mnO2`, `Cl`, `NO3`, `NH4`, `oPO4`, `PO4`, `Chla`), and test generalization of
model to data that have not been used for training.


4. **Cross-validation**: in class we talked about how to use cross-validation (CV) to estimate the "test error". In $k$-fold CV, each of $k$ equally sized random partitions of the data (chunks) are used in a heldout set (called validation set or test set). After $k$ runs, we average the held-out error as our final estimate of the validation error.  For this part, we will run cross-validation on only a single model, as a way to estimate our test error for future predictions (we are not using it here for model selection since we are considering only one model).  Perform 5-fold cross-validation on this model to estimate the (average) test error.
    
    #. \label{chunkids} First randomly partition data into 5 equal sized
    chunks. 

        Hint: a simple way to randomly assign each observation to a chunk is to
        do the following. First, use `cut(..., label=FALSE)` to divide
        observation ids (1, 2, \dots ) into equal numbers of chunk ids. Then,
        randomize output of `cut()`by using `sample()`.

```{r 4a) Partitioning Data into 5 chunks}
algae.folds <- cut(1:nrow(algae.med), breaks = 5, labels = FALSE) %>% sample()
algae.folds
```


    #. Perform 5-fold cross-validation with training error and validation
    errors of each chunk determined from \eqref{chunkids}. 

        Since same computation is repeated 5 times, we can define the following
        function for simplicity.

```{r cvtemplate,indent=indent2}
do.chunk <- function(chunkid, chunkdef, dat){  # function argument
  
    train = (chunkdef != chunkid)
    # print(paste("Train:", train))
    Xtr = dat[train,1:11]  # get training set
    Ytr = dat[train,12]  # get true response values in trainig set
    
    Xvl = dat[!train,1:11]  # get validation set
    Yvl = dat[!train,12]  # get true response values in validation set

    lm.a1 <- lm(a1~., data = dat[train,1:12])
    predYtr = predict(lm.a1)  # predict training values
    predYvl = predict(lm.a1, Xvl)  # predict validation values

    data.frame(fold = chunkid,
               # compute and store training error
               train.error = mean((predYtr - Ytr$a1)^2),
                # compute and store test error
               val.error = mean((predYvl - Yvl$a1)^2))  
}
```
        
        First argument `chunkid` indicates which chunk to use as validation set
        (one of 1:5). Second argument `chunkdef` is chunk assignments from
        \eqref{chunkids}. Third argument `dat` will be `algae.med` dataset.
        
        In order to repeatedly call `do.chunk()` for each value of `chunkid`,
        use functions `lapply()` or `ldply()`. Note that `chunkdef` and `dat`
        should be passed in as optional arguments (refer to help pages).

        Write the code and print out the `train.error` and `val.error` five times (e.g. for each chunk).

```{r}
n = 5
CV <- plyr::ldply(1:n, do.chunk, folds, algae.med)
CV
```

      
5. BAD PROBLEM (test error is not similar to validation error) **Test error on additional data**: now imagine that you actually get _new_ data that wasn't available when you first fit the model.

    #. Additional data can be found in the file `algaeTest.txt`.

```{r real,indent=indent2,message=F,warning=F}
algae.Test <- read_table2('../algae_data/algaeTest.txt',
                    col_names=c('season','size','speed','mxPH','mnO2','Cl','NO3',
                                'NH4','oPO4','PO4','Chla','a1'),
                    na=c('XXXXXXX'))
```
       
        This data was not used to train the model and was not (e.g. wasn't used in the CV procedure to estimate the test error).  We can get a more accurate measure of true test error by evaluating the model fit on this held out set of data.  Using the same linear regression model from part 4 (fit to all of the training data), calculate the "true" test error of your predictions based on the newly collected measurements in `algaeTest.txt`.  Is this roughly what you expected based on the CV estimated test error from part 4? 
       


**_Cross Validation (CV) for Model Selection_**

In this problem, we will be exploring a dataset of wages from a group of 3000 workers.  The goal in this part is to identify a relationship between wages and age.  

6.  First, install the `ISLR` package, which includes many of the datasets used in the ISLR textbook. Look at the variables defined in the `Wage` dataset.  We will be using the `wage` and `age` variables for this problem.  

```{r islr_install, indent=indent1, message=F, eval=FALSE, warning=FALSE}
library(ISLR)
head(Wage)
```


    #.  Plot wages as a function of age using `ggplot`.  Your plot should include the datapoints (`geom_point()`) as well as a smooth fit to the data (`geom_smooth()`).  Based on your visualization, what is the general pattern of wages as a function of age? Does this match what you expect?


    #.  In this part of the problem, we will find a polynomial function of age that best fits the wage data.  For each polynomial function between $p=0, 1, 2, ... 10$:
    
        #.  Fit a linear regression to predict wages as a function of $age$, $age^2$, ... $age^p$ (you should include an intercept as well).  Note that $p=0$ model is an "intercept-only" model.
        #.  Use 5-fold cross validation to estimate the test error for this model. Save both the test error and the training error.
        
```{r dowagechunks indent = indent2}
do.wage.chunk <- function(chunkid, chunkdef, dat, i){  # function argument
  
    train = (chunkdef != chunkid)
    
    Xtr = dat[train, -11]  # get training set
    Ytr = dat[train, 11]  # get true response values in trainig set
    
    Xvl = dat[!train, -11]  # get validation set
    Yvl = dat[!train, 11]  # get true response values in validation set
    
    model <- if(i==0) lm(wage ~ 1, data = dat[train,]) else lm(wage ~ poly(age, i), data = dat[train,])
    
    predYtr = predict(model)  # predict training values
    predYvl = predict(model, Xvl)  # predict validation values
    
    data.frame(fold = chunkid,
               # compute and store training error
               train.error = mean((predYtr - Ytr)^2),
                # compute and store test error
               val.error = mean((predYvl - Yvl)^2))  
}
```

```{r 6c) CV for model selection, indent = indent2}
n = 5
p = 10

test_error <- NA
train_error <- NA

wage.folds <- cut(1:nrow(Wage), breaks = n, labels = FALSE) %>% sample()

for(i in 0:p){
  wage.CV <- 1:n %>% plyr::ldply(do.wage.chunk, wage.folds, Wage, i)
  test_error[i] <-  mean(wage.CV$val.error)
  train_error[i] <-  mean(wage.CV$train.error)
}

test_error
train_error
```
        

---------------

7. __(231 Only)__ **The bias-variance tradeoff**.  Prove that the mean squared error can be decomposed into the variance plus bias squared.  That is, who $E[(\hat \theta - \theta)^2] = \text{Var}(\hat \theta)  + \text{Bias}(\hat \theta )^2$ where $\text{Bias}(\hat \theta) = E[\hat \theta - \theta]$.  Here $\hat \theta$ is an estimator (a random variable) of the fixed unknown constant $\theta$.  Hint: reogranize terms in the MSE by adding and subtracting $E[\hat \theta]$.



8. __(231 Only)__ As we discussed in class, distance metrics satisfy the following properties: 

- _Positivity_:

    * $d(x,y)\geq 0$
    * $d(x,y) = 0$ only if $x=y$

- _Symmetry_:
    * $d(x,y) = d(y,x)$ for all $x$ and $y$

- _Triangle Inequality_:
    * $d(x,z) \leq d(x,y) + d(y,z)$ for  $x,\ y,\text{ and } z$

    Show that the following measures are distance metrics by showing the above properties hold:

    #. $d(x,y) = \|x-y\|_2$


    #. $d(x,y) = \|x-y\|_\infty$
    

